---
title: "Data types and structure"
author: "Gustavo Bertrán"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data types and structure

1.  What are the different data types in R?
2.  What are the different data structures in R?
3.  How do I access data within the various data structures?

# Everything in R is an object

## `myobject<-10`

| myobject |         \<-         |               10               |
|:--------:|:-------------------:|:------------------------------:|
|  object  | assignment operator | what is stored in the operator |

# Data types in R

-   `class()` : what kind of object is it?
-   `length()` : how long is it? What about two dimensional objects?
-   `attributes()` : does it have any metadata?
-   `str()` :compactly display data

#Practice data

## Observe data

```{r}
library(datasets)
data("mtcars")
head(mtcars, 3)
print(head(mtcars))
class(mtcars)
length(mtcars) #number of columns
nrow(mtcars)   #number of rows
ncol(mtcars)   #number of columns
attributes(mtcars) #all of the information above
str(mtcars)  #strings, organizes data by column and the information each column has
summary(mtcars) #Gives central tendency measurements
```

## Manipulate Data

```{r}
four <- 4
class(four)

```

# Data structures in R

-   R has 6 basic data structures
    1.  Vector
    2.  Matrix
    3.  Raw(rearely used)
    4.  Data frame
    5.  List
    6.  Factor

## One Dimensional Structures

1.  Vector
2.  List

### Vector

-   Collection of elements that are mostly commonly of mode character, logical, integer or numeric.

-   `vector <- c(x,y,z)`

    -   R by default will classify it as a logical argument

    ```{r}
    vector()
    x <- c(1,2,3)
    x
    class(x)
    z <- c("Natally", "Natalia", "Ariana")
    class(z)
    #para agregar un elemento al vector utiliza
    z <- c(z,"Alan")
    z
    z <- sort(c(z,"Diana"))
    z
    z <- sort(z, decreasing = TRUE)
    z
    y <- c(x,z)
    class(y)
    y

    ```

    ### Coercion

-   If more than one type is included in a `c()` function, R silently coerces (forces) the vector to be of one type

    -   The type of element will focllow this heirarchy:

        -   logical\<integer\<numeric\<complex\<character \### Create your own vector

        ```{r}
        a <- seq(1:50) # provides a sequence of numbers
        seq(1, 50, by = 2)
        seq(from = 2, to = 20, by = 0.2)
         a[40] #shows the value in pos. 40
         z[2]
         z1 <- z[-2]
         z1
         
        ```

```{r}
series <- 0:100
series
sample(series,1) #random number from series
sample(series, 5, replace = TRUE) #values may repeat from random selection


```

## Missing Data

-   In R, missing data is represented by the symbol `NA` (not available)

```{r}
v <- c(1,2,NA,4,5)
mean(v)
mean(v,na.rm = TRUE)

is.na(v)

```

# Two Dimensional Structures

1.  Matrix
2.  Data frames

## Matrix

-   All columns in a matrix must have the same mode (numeric) character, etc.) and the same length

```{r}
m <- matrix(1:20,nrow=4,ncol=5) #Standard matrix organization
m
n <- matrix(1:20, nrow=4, ncol=5, byrow=FALSE)
n

dim(m) #observe dimensions of your matrix

j <- c(1,2,3,4,5)
k <- c(6,7,8,9,10)
l <- c(11,12,13,14,15)
matrix <- cbind(j,k,l)
dim(matrix)

cor(matrix)

```

## Data frames

-   Data frames are tabular data objects
-   Unlike a matrix, in data frames each column can contain different modes of data
-   We can create a data frame using the `data.frame()` function.

```{r}
df <-  data.frame(
  Name= c("Yoda", "R2_D2", "Chewbacca", "Obi-Wan Kenobi", "Luke Skywalker"),
  Age = c(900, 36, 235, 57, 53),
  Weight = c(130, 180, 150, 170, 160)
  )
df
```

# Subsetting Data Frames

-   You can access specific elements of data frame using square brackets `[]`
-   `$`operator can also be used to access specific columns by name

```{r}
df[1,]
df[,2]
df[1:3, 1]
subset_df <- df
subset(df, Age > 100 & Weight < 200)
subset
```

-   Add a column

```{r}
df$height <- c(66, 109, 228, 182, 172)
df$Jedi <- c(TRUE, FALSE, FALSE, TRUE, TRUE)
df
```

-   Add a row

```{r}
new_row <- data.frame(Name = "Darth Vader",
                      Age = 45,
                      Weight = 120,
                      height = 292,
                      Jedi = FALSE)
df1 <- rbind(df, new_row)
df1

```

-   changing names of a column

```{r}
colnames(df) <- c('character', 'Age (yrs)', 'Weight (kgs)', 'Height (m)', 'Jedi' )
df
str(df)

```

-   Order Data Frame by a specific column

```{r}
df2 <- data.frame(
  "character"= c("Yoda", "R2_D2", "Chewbacca", "Obi-Wan Kenobi", "Luke Skywalker", "Darth Vader"),
  Planet = c("Dagoba", "Tatooine", "Kashyyyk", "Naboo", "Stewjon", "Tatooine")
  )


merged_df <- merge(df, df2, by = 'character')
merged_df


```

```{r}
codon_usage <- read.csv("codon_usage.csv")
head(codon_usage)

str(codon_usage)

codon_usage[1]
viral <- codon_usage[codon_usage$Kingdowm == "vrl", ]
viral_bacteria <- codon_usage[codon_usage$Kingdom %in% c("vrl", "bct"),]
```

## Converting between data structures

-   We can convert between different data structures using functions
