---
title: 'Data Manipulation : dplyr'
author: "Gustavo Bertrán"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Wrangling with `dplyr`
## Rows
- `filter()`: keep rows matching conditions
- `arrange()` : reorder rows
- `slice()`   : select rows by position
## Columns
- `select()`: pick column by name or positions
- `rename()`: rename columns
- `mutate`  : create new columns
## Groups
- `group_by()` : define groups
- `summarize()`: collapse data within groups

## `%>%`: "then" function 

# `dplyr`: the pipe ` %>% `
- All dplyr funcitons take a data frame as the first argument.
- Without the pipe, you'd need nested calls or temporary objects.
-  The pipe can be read as "then" 
- It passes the result of one function into the nedxt.
- This makes your code:
  - Easier to read
  - Easier to write
  - Easier to debug
# Example dataset
```{r}
library(dplyr)
data("starwars")
head(starwars)


```
```{r, include=FALSE}
str(starwars
    )
```

# `dplyr`:`filter()`
- `filter()` selects a subset of rows in a data frame or tibble
- **The first argument is always the data frame**
- ** The second argument (and subsequent) are conditions on variables, keeping only rows where the condition is TRUE

```{r}
starwars %>% 
  filter(species == "Droid")
# in base r it would look like:
subset(starwars, species == "Droid")

#another example
starwars %>% 
  filter(skin_color=="light",
         eye_color=="brown" &
           hair_color=="black")

```

# `dplyr` : `arrange()`
- `arrange()` reorders the rows of a data frame or tibble
- The first argument is the data frame, followed by one or more column names (or expressions) to order by
- If multiple columns are provided, later columns are used to break ties in earlier ones

```{r}
starwars %>% 
  arrange(height, mass)


starwars %>% 
  arrange(desc(height))
```

# `dplyr`: `slice()`
- selects rows by their position
- Comes with several helper functions for common use cases:
  - `slice_head()`: select the first n rows
  - `slice_tail()`: select the last n rows

```{r}
starwars %>% 
  slice(5:10)

starwars %>% 
  slice_head(n=8)

starwars %>% 
  slice_tail(n=5)

```


# `dplyr`: `select()`
- Makes it easy to extract specific columns by name( or whith helper functions)

```{r}
starwars %>% 
  select(name, hair_color, skin_color, eye_color) %>% 
  slice_head(n=10)

```
- Use `!` to deselct (drop) columns:
```{r}
starwars %>% 
  select(!hair_color,! eye_color)
```
- We can use `-`to extract all except particular columns
```{r}
starwars %>% 
  select(-name, -skin_color)
```
- We can also specify column positions, or combine column positions and names
```{r}
starwars %>% 
  select(1,2, species) %>% 
  filter(species=="Droid") %>% 
  summarise(species)
```
#`dplyr`: `select()` with `contains()`
- To select only columns with names containing the letter "D"
```{r}
starwars %>% 
  select(contains('c'))
```
#`dplyr`: `select()` with `starts_with()`
- `starts_with()` and `ends_with()` offer more pecificity for `select()`
```{r}
starwars %>% 
  select(starts_with('n'))
```

# `dplyr`:`rename()`
- changes column names wihtout altering the data
```{r}

#First the name you want, second the current name
starwars %>% 
  rename(Character = name)


```

# `dplyr`:`mutate()`
- Add new columns that are function of existing ones
```{r}
newstarwars <- starwars %>% 
  mutate(height_m = height / 180)
newstarwars %>% #order of columns 
  select(height_m, height, everything())
```
- The `.keep` argument controls which columnsare kept in the output
```{r}
newstarwars %>% 
  mutate(height_m = height/100,
         .keep = "none")
```

#`dplyr`:`mutate()` with `ifelse()`
- Return:
  - One valuse **if** the condition is "TRUE"
  - Another value **if** the condition is "FALSE"
- Syntax: `ifelse(condition, value_if_true, value_if_false)`
```{r}
starwars %>% 
  mutate(height_cat = ifelse(height > 100, "tall", "small")) %>% 
  select (height, height_cat, everything())

```
#`dplyr`: `mutate()` with `ifelse()` and `ggplot2`
- After creating a new variable, we can quickly visualize it with ggplot2
```{r}
library(ggplot2)
starwars %>% 
  mutate(height_cat = ifelse(height > 100, "tall", "small")) %>% 
  ggplot(aes(x=height, file=height_cat)) +
  geom_histogram()+
  labs(x='algo', y='otro')


```

#`dplyr`:`summarise()`,`mean()`
- `summarise()` collapses a data frame to a single row( or one row per group if grouped)>
- Useful for computing summary statistics:
- `na.rm=TURE` ignores missing values in calculations
```{r}
starwars %>% 
  summarise(height_cm = mean(height, na.rm=TRUE))
```
#`dplyr`:`summarise()`,`mean()`,`min()`,`max()`
```{r}
starwars %>% 
  summarise(mean=mean(height, na.rm=T),
            min=min(height, na.rm=T),
            max=max(height, na.rm=T))
```

#`dplyr`: `group_by()`, `summarise()`
- `group_by()` defines groups for group-wise operations
- `summarise()` then computes summary satistics for each group
```{r}
starwars %>% 
  group_by(species) %>% 
  summarise(mean_height = mean(height, na.rm=T),
            sd_height= sd(height, na.rm=T)
            )
```

```{r}
starwars %>% 
  group_by(homeworld, gender) %>%
  summarise(mean_height_by_gender = mean(height, na.rm=T),
            sd_height_by_gender= sd(height, na.rm=T))
  
  
```
# `dplyr`: `count()`
- takes the name of one or more columns that contain the groups we are interested in, and we can optionally sort in order
```{r}
starwars %>% 
  group_by(species) %>% 
  count(homeworld, sort = TRUE)
```

#`dplyr`: `sample_n()`
- randomly selects a specified number of rows from a data frame
- Useful for exploring a subset of a large dataset
```{r}
starwars %>% 
  sample_n(10)

starwars %>% 
  sample_n(10,
           replace=T
           )
```

#`dplyr` recap
- All verbs in `dplyr` share a consisten grammar:
  - The first argument is always a data.frame (or tibble)
  - The next arguments describe what to do with the columns
  - You can refer to columns directly (no `$` needed)
  - The result is always a new data frame (tibble)
  
  msleep %>%
  filter(awake >= 12)


msleep %>%
  mutate(percent_day_awake = (awake / 24) * 100) %>%
  select(percent_day_awake, everything())
  
  
  storms %>%
  group_by(year) %>%
  summarise(num_storms = n())
  
  
  storms %>%
  filter(wind > 50) %>%
  arrange(desc(wind))
  
  
  
  lizards %>%
  group_by(Site, Diameter) %>%
  summarise(total_grahami = sum(grahami)) %>%
  arrange(desc(total_grahami))
  
  
  lizards %>%
  group_by(Time) %>%
  summarise(avg_opalinus = mean(opalinus)) %>%
  arrange(desc(avg_opalinus))

