---
title: "Practice"
author: "Gustavo Bertrán"
date: "2025-09-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries
```{r}

library(readxl)
library(dplyr)

```

### Practice data manupulation using Base R
The data in cereal.csv includes nutrition information collected from breakfast cereals. The data, which were provided by (Lock:2020?), were obtained from http://nutritionresource.com/.

- The attributes included in the data are:
  - name: Brand name of cereal
  - company: Manufacturer (G=General Mills, K=Kellog’s or Q=Quaker)
  - serving: Serving size (in cups)
  - calories: Calories (per cup)
  - fat: Fat (grams per cup)
  - sodium: Sodium (mg per cup)
  - carbs: Carbohydrates (grams per cup)
  - fiber: Dietary Fiber (grams per cup)
  - sugars: Sugars (grams per cup)
  - protein: Protein (grams per cup)

```{r, message=FALSE}
library(readxl)
cereal <- read_excel("cereal.xlsx")
head(cereal)
```

Five question ():
1. Extract rows 1 to 10 from the cereal data frame. (Extraiga las filas 1 a 10 del marco de datos cereal.)
```{r}
cereal_extracted <- subset(cereal[1:10,])
cereal_extracted
```

2. Create a new data frame called Kelloggs that only contains rows where company == “K”. (Cree un nuevo marco de datos llamado Kelloggs que solo contenga las filas donde company == “K”.)  
```{r}
df <- data.frame(subset(cereal, company == "K"))
df

```


3. From the cereal data, extract the names of all cereals that have more than 10 grams of sugar. (Del conjunto de datos cereal, extraiga los nombres de todos los cereales que tienen más de 10 gramos de azúcar.)  
```{r}
lowsugar <- cereal$name[cereal$sugars > 10]
lowsugar
dim(lowsugar)
```


4. Find the mean calories of all cereals in the dataset. (Encuentre las calorías promedio de todos los cereales en el conjunto de datos.)  
```{r}
totalcal <- cereal$calories
avgcal <- mean(totalcal)
avgcal
```


5. Count how many cereals have fiber greater than or equal to 2 grams (Cuente cuántos cereales tienen fibra mayor o igual a 2 gramos.).  

```{r}
highfiber <- subset(cereal, fiber >= "2")
highfiber
```



English In the cereal dataset, the first ten rows include cereals such as [names of some cereals from the first 10 rows], showing a range of values for calories, sugar, and fiber. Filtering for cereals from the company Kellogg’s gives a subset containing brands like [examples of Kellogg’s cereals], while focusing on cereals with more than 10 grams of sugar identifies products such as [names of high-sugar cereals]. The mean calories of all cereals in the dataset is approximately [mean calorie value], providing a general idea of the energy content of these products. Finally, counting cereals with at least 2 grams of fiber shows that [number of cereals with ≥2g of fiber] cereals meet a minimum dietary fiber content. This analysis helps to better understand the nutritional characteristics of the different cereals in the dataset.