---
title: "ANOVA"
author: "Gustavo Bertrán"
date: "2025-10-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("~/Intro R/Project R/R-Studio-Workarounds/assignments")
library(ggplot2)
library(car)
library(dplyr)
```


```{r}
library(readr)
tyre <- read_csv("tyre.csv")
head(tyre)
```


```{r}
p <- ggplot(tyre, aes(x = Brands, y = Mileage)) + 
  geom_boxplot(aes(fill=Brands)) +
  geom_jitter() +
  #Labels
  labs(title= "Rendimiento de Gomas en PuertoRico", x="Gomas", y="Kilometers") +
#Themes
  theme_linedraw() +
  theme(plot.title = element_text(size = 18,
                                  face = "bold",
                                  hjust = 0.5,
                                  color = "cornflowerblue")) + 
  theme(axis.title.y = element_text(size =18,
                                    face ="bold",
                                    color="lightblue"),
        axis.text.y = element_text(size = 16,
                                   face = "plain",
                                   color = "black")) +
  theme(axis.title.x = element_text(size = 18,
                                    face = "bold",
                                    color = "blue"),
        axis.text.x = element_text(size = 16,
                                   face = "plain",
                                   color = "black")) +
  theme(legend.title = element_text(size = 18,
                                    face = "italic",
                                    color = "black")) 
        
  
  

p

```

#Assumption checks
```{r}
mod <- aov(Mileage ~ Brands, data = tyre)
summary(mod)

residuals_mod <- residuals(mod)

shapiro.test(residuals_mod)

leveneTest(Mileage ~ Brands, data = tyre)
#high p value = equal variance


```

```{r}
post_test <- TukeyHSD(mod)
post_test
```


```{r}

#ggsave(filename="plot.png",
#plot = p,
#width=8,
#height=6,
#dpi=300)

```


# Graph analysis Description

The data analyzed was named tyre, which captured the mileage different wheel brands have. Through several test I looked for which brand was the most effective (more milage per brand) and if it was significant. I performed a one way anova test and tested for normality of its residuals through the Shapiro test (pvalue = 0.78). Afterwards I tested for Homogeneity of Variance through Levene's Test (pvalue = 0.56). After confirming equal variance and normal distribution of the residuals I performed a Tukey test to determine how significant where the differences in Milage. In sumarry Falken showed superiority over every other brand tested(Falken had a p value lower thatn 0.002 in every comparison.[the value was near 0 when compared to Bridgestone]). In contrast, Bridgestone performed the worst when compared. Between both fell Apollo and CEAT with no significant difference within each other. A final ranking was determined to be: 1. Falken 2. CEAT/Apollo 3. Bridgestone. To better visualize the results; using `ggplot` I graphed a boxplot with a jitter plot. Text size and color was adjusted for the title, axis and legend for better visualization. 
