---
title: "ggplot"
author: "Gustavo Bertrán"
date: "2025-10-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plots in Base R
- There are many types of statistical plots, but only five essential ones for beginners:
  - Scatter plots: show relationships among numberical variables.
  - Line graphs: show change over time.
  - Histograms: show data distributions
  - Boxplots: show between-group and within-group variation
  - Baar plots: show summary statistics (e.g., counts, means, or proportions)
  
## Thinking like a statistician
- Quantitative Variables: a measurable amount
  1. Continous variable: may assyme any imaginable value within a certain rage. Can (theoretically) have an infinite number of values
    - Weights, Heights
  2. Discrete Variables; countable as integers (whole numbers). No values between two adjacent values are permissible
    - Number of bicycles sold in a day.
    
- Qualitative Variables: descriptive characteristic assignable to a category.
  1. Nominal Variables: measurements fall into a particular class or categorty with no order implied.
    -sex(male or female), color(red, blue, green)
  2. Ordinal Variables: a ranking scale where order between categories is implied.
    - Likert scale(strongly agree, agree, neutral, disagree, strongly disagree)
    3. Interval (ratio) Variables: use a quantitative measurement to assign a specific qualitative category(these are still ordinal)
    - Temperature( cold, warm, hot), age(young, middle-aged, old)
    
## Plots in Base R: Barplots
- In a bar chart, calues are represented by length of bars, each corresponding to a measured group. 
  - Bar charts can be vertical or horizontal: 
    - Vertical bar charts are often called column charts.
    - Horizontal bar charts are ideal wehen:
      - You have many bars to plot, or labels need extra space to be legible.
    
## Dataset: DNase(built-in in R)
- We will use `barplot()`function to create a barplot of the mean optical density for each substrate concentration  

#Load dataset
```{r}
data("DNase")
head(DNase)
```
- Variables
  - `conc` -> substrate concenctration (in mg/mL).
  - `density` -> optical density, representing the experimentally measured enzymatic activity.
  - `Run` -> experimental run or replicate number (factor with 11 levels).
  
## `barplot()` function
- R language uses the the function `barplot()` to create bar charts. Here, both vertical and Horizontal bars can be drawn.  
- Syntax: `barplot(H, names.arg, xlab, ylab, main, col)` 
  -`H`  vector or matrix containing numeric values which are used in bar chart
  - `names.arg` vector of names appearing under each bar in bar chart
  - `xlab`  label for x axis in bar chart
  - `ylab` label for y axis in bar chart
  - `main`  title of the bar chart
  - `col` used to give colors to the bars in the graphsd
  
  
### # 1 Inspect Data
```{r}
head(DNase, 3)

# Variables:
# conc = concentration of the substrate
# density = optical density (enzyme activity)
# Run = experimental replicate
```

- We use `aggregate()` to summarize the 11 experimental runs and visualize the average enzymatic activity at each concentration.

### # 2 Calculate mean density per concentration

```{r}
mean_density <-  aggregate(density ~ conc, data = DNase, FUN = mean)
  
  
``` 

### # 3 Create barplot
```{r}
barplot(
  height = mean_density$density,
  names.arg = mean_density$conc,
  col = blues9,
  main = "Mean DNase I Activity by Concentration",
  xlab = "Concentration (mg/mL)",
  ylab = "Mean Optical Density",
  las = 2,
  cex.names = 0.54
)
```

- Simple Horizontal Bar Plot
```{r}
barplot(
  height = mean_density$density,
  names.arg = mean_density$conc,
  col = blues9,
  main = "Mean DNase I Activity by Concentration",
  xlab = "Concentration (mg/mL)",
  ylab = "Mean Optical Density",
  las = 2,
  cex.names = 0.54,
  horiz = TRUE
)
```

## Dataset: coronary
- The dataset contains the total cholesterol level, their individual characteristics and intervention groups in a hypothetical clinical trial. The dataset contains 200 observations for nine variables:
```{r}
setwd("~/Intro R/Project R/R-Studio-Workarounds/assignments")
library(readxl)
coronary <- read_excel("coronary.xlsx")
head(coronary, 3)

```
- The dataset contains the total cholesterol level, their individual characteristics and intervention groups in a hypothetical clinical trial. The dataset contains 200 observations for nine variables:
- Variables:
  - id: subject's id
  - cad: Coronary artery disease status (categorical) {no cad, cad}
  -sbp: Systolic blood pressure in mmHg (numerical)
  -dbp: Diastolic blood pressure in mmHg (numerical)
  -chol: Total cholesterol level in mmol/L (numerical)
  -age: Age in years (numerical)
  -bmi: Body mass index (numerical)
  -race: Race of the subjects (categorical) {malay, chinese, indian}.
  -gender: Gender of the subjects (categorical){woman, man}

### Create a line chart
```{r}
plot(
  coronary$dbp ~ coronary$chol,
  type = "p",
  col = "blue",
  lwd = 2,
  xlab = "Total Cholesterol (mol/L)",
  ylab = "Diastolic Blood Pressure (mmHg)",
  main = "Relationship between Cholesterol and Diastolic BP",
  pch = 15
)

?plot
```

## Spearman Correlation: chol vs dpb
- The Spearman correlation evaluates the monotonic (rank-based) relationship between cholesterol and diastolic blood pressure.

```{r}
spearman_result <- cor.test(
  coronary$chol,
  coronary$dbp,
  method = "spearman",
  exact = FALSE
)

spearman_result
?cor.test
```

- Spearman's rho (p) indicates the strength and direction of the monotonic relationship.
- p-value assesses wether this relationship is statistically significant
  - p > 0 -> higher cholesterol toends to accompany higher diastolic BP
  - p < 0 -> higher cholesterol tends to accompany lower diastolic BP
  
  
## Test for Normality

```{r}
shapiro.test(coronary$chol)

shapiro.test(coronary$dbp)



```

## Visualization

```{r}
plot(
  coronary$dbp ~ coronary$chol,
  type = "p",
  col = "blue",
  lwd = 2,
  xlab = "Total Cholesterol (mol/L)",
  ylab = "Diastolic Blood Pressure (mmHg)",
  main = "Relationship between Cholesterol and Diastolic BP",
  pch = 15
)
abline(lm(dbp ~ chol, data = coronary),col = "red", lwd = 2, lty =2)
```

## Line Charts
- Line charts show changes in value across continous measurements, such as over time.
- Movement of the line up or down highlights positive or negative changes.
- Line charts can expose overall trends and helmp make predictions or projections
- Multiple line charts can also give rise to related charts like sparklines or ridgeline plots.

## plot() function
```{r}
coronary <- coronary[order(coronary$age), ]

plot(coronary$age, coronary$chol,
     type = "l",
     col = "blue",
     lwd = 2,
     xlab = "Age (years)",
     ylab = "Cholesterol (mmol/L)",
     main = "Cholesterol vs Age"
)
```

## Histograms
- Histograms are used to visualize distribution of a single continous variable
- They show how often data values fall within specific intervals 
- Useful for identifying the shape of the data and detecting outliers or spread in the dataset
- in R, histograms can be created using the `hist()` function, which allows customization of the number of bins, colors, and labels


```{r}
hist(coronary$chol,
     main = "Distribution of Cholesterol",
     xlab = "Cholesterol (mmol/L)",
     col = "lightblue",
     border = "white",
     breaks = 40
     )
```


## Boxplots
- Boxplots are used to visualize the distribution of a numeric variable and detect outliers
- They display:
  - Median (central line inside box)
  - Interquartile range (IQR) - the box spans from the 25th to the 75th 
  - Whiskers which extend up to Q1 -1.5 & Q3 +1.5 from the box
  - Points beyond whiskers represent potential outliers
  
```{r}
boxplot(coronary$chol,
        main = "Cholesterol Levels",
        ylab = "Cholesterol (mmol/L)",
        col = "lightgreen",
        border = "darkgreen"
        )
```


```{r}
boxplot(sbp ~ gender,
        data = coronary,
        main = "Systolic Blood Pressure by Age Group",
        xlab = "Age Group (years)",
        ylab = "Systolic BP (mmHg)",
        col = "lightblue",
        border = "darkblue"
        )
```

## Variance analysis
- We can use ANOVA to test wether mean systolic blood pressure (SBP) differs across age groups or genders.

```{r}
anova_model <- aov(sbp ~ gender, data = coronary)
summary(anova_model)

```

## Assumptions of ANOVA
- Normality- residuals should be approximately normally distributed. We check this using the Shapiro Test
```{r}
shapiro.test(residuals(anova_model))
```


