---
title: "Tables"
author: "Gustavo Bertrán"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 
# Tables  
- We can create a table by using `as.table()` function, first we create a table using matrix and then assign it to thid method to get the table format.

# Tables `knitr::kable()`  
- The `kable()` function is provided by the `knitr` package
- It creates and formats tables in Markdown HTMLor LaTex
- Partiularly useful in reports or documents where you need professional-looking tables.
Works directly with data frames and matrcies, making it easy to integrate into analysis and reporting workflows.
- By default, `kable()` produces a clean table, but it can be extended with packages like `kableExtra` for more advanced formatting

```{r}
library(gaawr2)
library(dplyr)
library(kableExtra)
library(knitr)
library(gt)
data(diabetes)
head(diabetes)
```

```{r}
kable(head(diabetes,5),bbooktabs = TRUE) %>% 
  kable_styling(font_size = 12) 
```

```{r}
kable(head(diabetes,5), booktabs = TRUE) %>% 
  kable_styling(font_size = 14,bootstrap_options = "striped", full_width = F)
```
```{r}
kable(head(diabetes[1:2],5), booktabs = TRUE) %>% 
  kable_styling(font_size = 14,bootstrap_options = "striped", full_width = F)
```


```{r}
kable(head(diabetes[1:2],5), booktabs = TRUE) %>% 
  kable_styling(font_size = 14, position = "right", bootstrap_options = "striped", full_width = F)
```

```{r}
kable(head(diabetes, 5), booktabs = TRUE) %>% 
  kable_styling(font_size = 14,bootstrap_options = "striped", full_width = F) %>% 
  add_header_above(c("Personal data" = 4, "Profile" = 10))
```
```{r}
kable(diabetes, digits = 2, format = "html", row.names = TRUE) %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = T,
                font_size = 15) %>% 
  scroll_box(height = "300px")


```



```{r}
diabetes %>% 
  dplyr::select(ID, No_Pation, Gender, AGE) %>% gt_preview()
```

